Experimentally informed site-specific substitution models deepen phylogenetic estimates of the divergence of viral lineages

### 2018-02-16 manuscript/manuscript.tex
updates to Results section 2

Here we consider how purifying selection is handled by codon substitution models, which are the most accurate of the three classes (nucleotide, amino acid, and nucleotide) of phylogenetic substitution models in widespread use\jdbcomment{find a citation for this and replace word ``accurate'' with something else if appropriate. Perhaps this one or something cited within \url{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4620419/}}.
Codon substitution models distinguish between two types of substitutions: synonymous and nonsynonymous.
The relative rate of these substitutions is referred to as dN/dS or $\omega$.
In their simplest form, codon substitution models fit a single $\omega$ that represents the gene-wide average selection on nonsynonymous mutations relative to synonymous ones.
Here we will use such substitution models in the form proposed by \citet{goldman1994codon}, which we denote GY94.
Using the classification scheme of \citet{yang2000codon}, when these models have a single gene-wide $\omega$ they are called GY94 M0.
This gene-wide $\omega$ is usually $<1$\jdbcomment{add citation, most of the site-specific $\omega$ papers will talk about how gene-wide is not sensitive, so we can cite some of those}, and crudely represents the fact that many amino-acid substitutions are under purifying selection.

@@

A model formulation that accounts for the fact that purifying selection depends on the specific amino-acid mutation is provided by so-called ``mutation-selection'' models~\citep{halpern1998evolutionary}.
These models by explicitly define a different set of amino-acid preferences at each site in the protein. 
This more mechanistic formulation results in a site-specific stationary state (\ref{fig:model_feature}B). 
These models better capture the site-to-site variation in amino-acid composition that is an obvious features of real proteins, and generally better describe actual evolution than models with only rate variation~\citep{lartillot2004bayesian, le2008phylogenetic, rodrigue2010mutation,hilton2017phydms,bloom2014experimentally}.

However, the increased realism of mutation-selection models comes at the cost of an increased number of parameters. 
Codon substitution models with uniform stationary states have only a modest number of parameters that must be fit from the phylogenetic data.
For instance, a GY94 M5 model with the typical F3X4 stationary state\jdbcomment{add citation, maybe just GY94 paper?} has 12 parameters: two describing the shape of the gamma distribution over $\omega$, a transition-transversion rate, and 9 parameters describing nucleotide composition of the stationary state.
However, mutation-selection models must additionally specify 19 parameters defining the stationary state for \emph{each} site (there are 20 amino acids whose frequencies are constrained to sum to one).
This corresponds to $19\times L$ parameters for a protein of length $L$, or 9500 parameters for a 500-residue protein.
It is challenging to obtain values for these parameters without overfitting the data\skhcomment{cite rodrigue paper}.
Here we will primarily use experimentally informed codon models (ExpCM)~\citep{bloom2014experimentally, hilton2017phydms, bloom2017identification} which define values for these parameters \textit{a priori} from deep mutational scanning experiments~\citep{fowler2014deep} so they do not need to be fit from phylogenetic data.
Therefore, the number of remaining free parameters for an ExpCM are similar to a non-site-specific substitution model.
Alternative strategies of obtaining parameters for site-specific stationary states via Bayesian\skhcomment{cite} or maximum-likelihood estimation\skhcomment{cite} are discussed in the last section of the Results.

Importantly, these two strategies for modeling purifying selection are not mutually exclusive.
Mutation-selection models such as ExpCM can still incorporate an $\omega$ parameter, which now represents the relative rate of nonsynonymous to synonymous substitution \emph{after} accounting for the constraints due to the site-specific stationary state.
This $\omega$ parameter can be drawn from a distribution just like as for GY94-style models. 
We will denote ExpCM's that also include a distribution over $\omega$ as ExpCM+$\Gamma\omega$.
\ref{fig:model_feature}C shows the full spectrum of models that incorporate all combinations of gamma-distributed $\omega$ and site-specific stationary states.

@@

Given a single branch, a substitution model transforms sequence identity into branch length.
Under a molecular-clock assumption, this branch length is proportional to time.
The transformation from sequence identity to branch length is trivial and insensitive to the choice of substitution model when the sequence identity is high.
For instance, in the case when where there has only been one substitution, then the sequence identity will simply be $\frac{L - 1}{L}$ for a gene of $L$ sites, and even a simple exponential model~\citep{zuckerkandl1965} will correctly infer the short branch length of $1/L$ substitutions per site.
However, as more substitutions accumulate then it becomes progressively more likely for multiple changes to occur at the same site.
In this regime, the accuracy of the substitution model becomes critical for transforming the sequence identity into branch length.
Any time-homogenous substitution model predicts that after a very large number of substitutions, two related sequences will approach some asymptotic sequence identity.
For instance, if all 20 amino acids are equally likely in the stationary state, then this asymptotic sequence identity will be 0.05.
If the substitution model underestimates the asymptotic sequence divergence then it will also underestimate long branch lengths, since it will predict that long branches lead to more divergence lead than is actually the case.

\ref{fig:decay} shows how different substitution models predict sequence identity to decrease as a function of branch length for model parameters fit to a phylogeny of an H1 influenza hemagglutinin (HA).
The GY94 and GY94+$\Gamma\omega$ models each predict the same behavior for all sites, since they do not have any site-specific parameters.
Both of these models predict an asymptotic sequence divergence of \jdbcomment{?}, which is slightly higher than 0.05 since some of the 20 amino acids are favored by the models due to more redundant codons and biases towards certain nucleotides.
Intuitively, this asymptotic sequence identity seems low, since like many proteins HA has a highly conserved structure and function which imposes constraints that cause some sites to only sample a small subset of the 20 amino acids among all known HA homologs.

\jdbcomment{Then talk about we have a sense of these constraints from the DMS, and if we use them to put an $\omega_r$ it changes rate of approach of asymptote but not asymptote.
Then ExpCM with different asymptotes.
Then point about stationary state.}

@@

Modeling site-to-site rate variation by $\Gamma\omega$ affects the rate at which stationary state is approached but not the sequence composition at stationary state itself. 
For example, in \ref{fig:decay}, GY94+$\Gamma\omega$ (green) takes longer to reach the stationary state than GY94 (blue) but stationary state for each model is identical. 
Because the stationary state is invariant to multiplicative terms, increasingly complex modeling of $\omega$ will never fully recapitulate the difference in expected sequence composition between the ExpCM and the GY94 at constrained sites. 
We inferred an $\omega$ value from the ExpCM for each site \citep{spielman2015relationship} and applied this values to the GY94 model (\ref{fig:decay} yellow)
These site-specific values drastically decreased the amount of time it takes to reach stationary state at the constrained sites but still does not affect the stationary state. 
This means that no matter how ``well" a model accounts for site-to-site rate variation, it will underestimate long branches with a uniform stationary state.

>>>>

Here we consider how purifying selection is handled by codon substitution models, which are the most accurate of the three classes (nucleotide, amino acid, and nucleotide) of phylogenetic substitution models in widespread use\jdbcomment{find a citation for this and replace word ``accurate'' with something else if appropriate. Perhaps this one or something cited within \url{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4620419/}}.
Codon substitution models distinguish between two types of substitutions: synonymous and nonsynonymous.
The relative rate of these substitutions is referred to as dN/dS or $\omega$.
In their simplest form, codon substitution models fit a single $\omega$ that represents the gene-wide average selection on nonsynonymous mutations relative to synonymous ones.
Here we will use such substitution models in the form proposed by \citet{goldman1994codon}.
Using the classification scheme of \citet{yang2000codon}, when these models have a single gene-wide $\omega$ they are called M0.
We will refer to the M0 variant of the \citet{goldman1994codon} model simply as GY94.
The gene-wide $\omega$ is usually $<1$\jdbcomment{add citation, most of the site-specific $\omega$ papers will talk about how gene-wide is not sensitive, so we can cite some of those}, and crudely represents the fact that many amino-acid substitutions are under purifying selection.

@@

A model formulation that accounts for the fact that purifying selection depends on the specific amino-acid mutation is provided by so-called ``mutation-selection'' models~\citep{halpern1998evolutionary}.
These models by explicitly define a different set of amino-acid preferences at each site in the protein. 
This more mechanistic formulation results in a site-specific stationary state (\ref{fig:model_feature}B). 
These models capture the site-to-site variation in amino-acid composition that is an obvious features of real proteins, and generally better describe actual evolution than models with only rate variation~\citep{lartillot2004bayesian, le2008phylogenetic, rodrigue2010mutation,hilton2017phydms,bloom2014experimentally}.

However, the increased realism of mutation-selection models comes at the cost of an increased number of parameters. 
Codon substitution models with uniform stationary states have only a modest number of parameters that must be fit from the phylogenetic data.
For instance, a GY94 M5 model with the typical F3X4 stationary state\jdbcomment{add citation, maybe just GY94 paper?} has 12 parameters: two describing the shape of the gamma distribution over $\omega$, a transition-transversion rate, and 9 parameters describing nucleotide composition of the stationary state.
However, mutation-selection models must additionally specify 19 parameters defining the stationary state for \emph{each} site (there are 20 amino acids whose frequencies are constrained to sum to one).
This corresponds to $19\times L$ parameters for a protein of length $L$, or 9500 parameters for a 500-residue protein.
It is challenging to obtain values for these parameters without overfitting the data~\citep{rodrigue2013statistical}.
Here we will primarily use experimentally informed codon models (ExpCM)~\citep{bloom2014experimentally, hilton2017phydms, bloom2017identification} which define values for these parameters \textit{a priori} from deep mutational scanning experiments~\citep{fowler2014deep} so they do not need to be fit from phylogenetic data.
Therefore, the number of remaining free parameters for an ExpCM are similar to a non-site-specific substitution model.
Alternative strategies of obtaining parameters for site-specific stationary states via Bayesian\skhcomment{cite} or maximum-likelihood estimation\skhcomment{cite} are discussed in the last section of the Results.

Importantly, these two strategies for modeling purifying selection are not mutually exclusive.
Mutation-selection models such as ExpCM can still incorporate an $\omega$ parameter, which now represents the relative rate of nonsynonymous to synonymous substitution \emph{after} accounting for the constraints due to the site-specific stationary state.
This $\omega$ parameter for an ExpCM can be drawn from a distribution just like for GY94-style models. 
We will denote such models as ExpCM+$\Gamma\omega$.
\ref{fig:model_feature}C shows the full spectrum of models that incorporate all combinations of gamma-distributed $\omega$ and site-specific stationary states.

@@

Given a single branch, a substitution model transforms sequence identity into branch length.
Under a molecular-clock assumption, this branch length is proportional to time.
The transformation from sequence identity to branch length is trivial when the sequence identity is high.
For instance, when where there has only been one substitution, then the sequence identity will simply be $\frac{L - 1}{L}$ for a gene of $L$ sites, and even a simple exponential model~\citep{zuckerkandl1965} will correctly infer the short branch length of $1/L$ substitutions per site.
However, as more substitutions accumulate then it becomes progressively more likely for multiple changes to occur at the same site.
In this regime, the accuracy of the substitution model becomes critical for transforming the sequence identity into branch length.
Any time-homogenous substitution model predicts that after a very large number of substitutions, two related sequences will approach some asymptotic sequence identity.
For instance, if all 20 amino acids are equally likely in the stationary state, then this asymptotic sequence identity will be 0.05.
If the substitution model underestimates the asymptotic sequence divergence then it will also underestimate long branch lengths, since it will predict that long branches lead to more divergence lead than is actually the case.

\ref{fig:decay} shows how different substitution models predict sequence identity to decrease as a function of branch length for model parameters fit to a phylogeny of H1 influenza hemagglutinin (HA) genes.
The GY94 predicts the same behavior for all sites, since it does not have any site-specific parameters.
This model predicts an asymptotic sequence divergence of \jdbcomment{?}, which is slightly higher than 0.05 since some of the 20 amino acids are favored due to more redundant codons and biases towards certain nucleotides.
Intuitively, this asymptotic sequence identity of \jdbcomment{?} seems low, since like many proteins HA has a highly conserved structure and function which imposes constraints that cause some sites to only sample a small subset of the 20 amino acids among all known HA homologs.

Accounting for site-to-site rate variation in GY94 models affects the rate at which the asymptotic sequence identity is approached, but not the actual value of this asymptote. 
For instance, \ref{fig:decay} shows that the GY94+$\Gamma\omega$ model takes longer to reach the asymptote than GY94, but the asymptote for both models is identical. 
This fact holds true even if we use experimental measurements of HA's site-specific amino-acid preferences~\citep{doud2016accurate} to calculate a different $\omega_r$ value for each site using the method of \citet[\ref{eq:w_r};][]{spielman2015relationship}.
Specifically, this GY94+$\omega_r$ model predicts that different sites will approach the asymptote at different rates, but the asymptote is always the same (\ref{fig:decay}).
The invariance of the asymptotic sequence identity under different schemes for modeling $\omega$ is a fundamental feature of the mathematics of homogeneous substitution models.
The asymptotic sequence divergence is a function of only the stationary state of the substitution matrix, and the stationary state of a stochastic matrix is invariant with respect to multiplication by any non-zero number\jdbcomment{But this isn't true, because $\omega$ only multiplies some elements of transition matrix. I think it's probably something like if you multiple pairs of elements on equivalent positions across diagonal?}.
Therefore, different schemes for modeling $\omega$ can alter the time it takes to reach stationary state, but no matter how ``well" a model accounts for site-to-site variation in $\omega$ it will always have the same stationary state as a simple GY94 model.

\jdbcomment{In contrast, ExpCM models have different stationary states at each site...}

@@
